---
title: "Severe Weather Event Analysis"
author: "Karl Wenzel"
date: "April 8, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

Using data obtained from the U.S. National Oceanic and Atmospheric Administration's (NOAA) [storm database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2), along with the analysis described later in this report, it was determined that several strong patterns exist between the types of severe weather events vs. damage, injury and human death.  **With respect to population health:** wind and tornodo related events dominate all other categrories in terms of being the **most harmful**.  Regarding damage caused by weather events: floods, storms, rain and hail cause **the greatest economic consequences**, with wind and tornodoes being next beyond that.  It was also determined that different states had different results, and a pair of **sweet maps** are provided herein to explain these results in better detail.

The code and reference data files used to create this document are available on github at [https://github.com/KarlWenzel/NOAA-Storm-Analysis](https://github.com/KarlWenzel/NOAA-Storm-Analysis), and the data set was obtained from the NOAA storm database as listed earlier.

## Data Processing

```{r message=FALSE}
library(plyr)
library(ggplot2)
library(gridExtra)
library(maps)
library(RColorBrewer)
```

```{r data-processing}
RAW = read.csv("repdata-data-StormData.csv.bz2")

# We will store our processed data in STORMS
STORMS = data.frame(
  date =  strptime( RAW$BGN_DATE, format="%m/%d/%Y" ),
  state = RAW$STATE,
  evnt =  as.character( RAW$EVTYPE ),
  dead =  RAW$FATALITIES,
  hurt =  RAW$INJURIES,
  prop =  RAW$PROPDMG,
  pexp =  tolower( RAW$PROPDMGEXP ),
  crop =  RAW$CROPDMG,
  cexp =  tolower( RAW$CROPDMGEXP )
)

# Filter out extraneous data
row.filter = (STORMS$dead + STORMS$hurt + STORMS$prop + STORMS$crop) > 0
STORMS = STORMS[ row.filter, ]

# Store year as a number
STORMS$year = as.numeric( format(STORMS$date, "%Y") )

# Since all data prior to 1955 is only for tornados, this data is discarded to avoid skewing
#   see http://www.ncdc.noaa.gov/stormevents/details.jsp
STORMS = STORMS[ STORMS$year > 1955, ]

# Prepare exponent reference data; use to join a multiplier based on the exponent key
my.exp = c( 1:9, "h", "k", "m", "b" )
my.mul = c( 10^(1:9), 100, 1000, 1000000, 1000000000 )
PEXP = data.frame( pexp = my.exp, pmul = my.mul )
CEXP = data.frame( cexp = my.exp, cmul = my.mul )

# Join the multipliers by matching the exponent keys
STORMS = join( STORMS, PEXP, by="pexp", type="left", match="first" )
STORMS = join( STORMS, CEXP, by="cexp", type="left", match="first" )

# Any remaining unset conditions for the multipliers are set to 0
STORMS[ is.na(STORMS$pmul),]$pmul = 0
STORMS[ is.na(STORMS$cmul),]$cmul = 0

# Scale the property damage and crop damage values by their multipliers
STORMS$prop = STORMS$prop * STORMS$pmul
STORMS$crop = STORMS$crop * STORMS$cmul

# Prepare category reference data; this provides us with regex patterns 
CATEGORIES = read.csv("category-reference.csv", stringsAsFactors=FALSE)
print(CATEGORIES)

# Use the patterns to assign categories

STORMS$category = NA
for (i in 1:nrow(CATEGORIES)) {
  filter = is.na(STORMS$category) & grepl( CATEGORIES[i,]$patterns, STORMS$evnt )
  STORMS[filter,]$category = CATEGORIES[i,]$category
}

# Any unset categories are grouped into their own category; promote to factor
STORMS[ is.na(STORMS$category), ]$category = "Other"
STORMS$category = as.factor(STORMS$category)

# Remove the columns that are no longer necessary
STORMS$cexp = NULL
STORMS$pexp = NULL
STORMS$X = NULL
STORMS$pmul = NULL
STORMS$cmul = NULL

# We now have our tidy data set.
print( head(STORMS) )

# This is a good saving point if you are running my code.
#   It creates a 17.8 MB file as opposed to the 548 MB uncompressed raw file
write.csv(STORMS, "storms-tidy-dataset.csv")
```

## Results

```{r results}
# First of all, determine the basic results
BASIC.SUMMARY = ddply(STORMS, .( category), summarize, 
  sum.people = sum(dead) + sum(hurt),
  sum.money = sum(prop) + sum(crop)
)
print( BASIC.SUMMARY )

# Render 2 plots in one image.  Show total damages and injuries, by category

basic.categories = levels(BASIC.SUMMARY$category)
color.bins = length(basic.categories)
colors = brewer.pal(color.bins, "Paired")

p1 = ggplot( data=BASIC.SUMMARY, aes(category, sum.people/1000) )
p1 = p1 + geom_bar(stat="identity", fill=colors)
p1 = p1 + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p1 = p1 + xlab("Weather Event Category") + ylab("People Killed or Injured in Thousands")
p1 = p1 + ggtitle("People Hurt by Category")

p2 = ggplot( data=BASIC.SUMMARY, aes(category, sum.money/1000000000) )
p2 = p2 + geom_bar(stat="identity", fill=colors)
p2 = p2 + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p2 = p2 + xlab("Weather Event Category") + ylab("Damage in Billions of Dollars")
p2 = p2 + ggtitle("Damage by Category")

grid.arrange(p1, p2, ncol=2)

# Now lets break it down by state
SUMMARY = ddply(STORMS, .(state, category), summarize,
  sum.people = sum(dead) + sum(hurt),
  sum.money = sum(prop) + sum(crop)
)

# Determine the category with the greatest cause of money damage by state
MONEY = ddply(SUMMARY, .(state), function(x){x[which.max(x$sum.money),]})

# Load state reference data; we'll use this for matching to states on our map
STATES = read.csv("states.csv")

# Join the state reference data with our damage report by state
MONEY$abbrev = MONEY$state
STATES.MONEY = join( STATES, MONEY, by="abbrev", type="left", match="all" )

# Reduce the category factor to only those in use
STATES.MONEY$category = factor(STATES.MONEY$category)

# Prepare our color palatte and category names
category.names = levels(STATES.MONEY$category)
color.bins = length(category.names)
colors = brewer.pal(color.bins, "Set1")

# Draw Map
map("state", col = "gray80", fill = TRUE, lty = 0) #Base
map("state", col = colors[STATES.MONEY$category], fill = TRUE, lty = 0, add = TRUE) #States
map("state", col = "gray", lwd = 1.4, lty = 1, add = TRUE) #Borders
title("Greatest Cause of Damage in Dollars by State")
legend("bottomright", cex= 0.56, bty="n",
   title="Source of Damage",
   legend = category.names,
   fill = colors
)

# Determine the category with the greatest cause of death and injury by state
PEOPLE = ddply(SUMMARY, .(state), function(x){x[which.max(x$sum.people),]})

# Load state reference data; we'll use this for matching to states on our map
STATES = read.csv("states.csv")

# Join the state reference data with our injury report by state
PEOPLE$abbrev = PEOPLE$state
STATES.PEOPLE = join( STATES, PEOPLE, by="abbrev", type="left", match="all" )

# Reduce the category factor to only those in use
STATES.PEOPLE$category = factor(STATES.PEOPLE$category)

# Prepare our color palatte and category names
category.names = levels(STATES.PEOPLE$category)
color.bins = length(category.names)
colors = brewer.pal(color.bins, "Set1")

# Draw Map
map("state", col = "gray80", fill = TRUE, lty = 0) #Base
map("state", col = colors[STATES.PEOPLE$category], fill = TRUE, lty = 0, add = TRUE) #States
map("state", col = "gray", lwd = 1.4, lty = 1, add = TRUE) #Borders
title("Greatest Cause of Death and Injury by State")
legend("bottomright", cex= 0.56, bty="n",
   title="Source of Damage",
   legend = category.names,
   fill = colors
)


```

## The End

